---
title: 'Outliers and Normalisation: mito, miRNA, bulk RNA for skeletal musc'
author: "MSoria"
date: "2025-21-Jul"
output:
  html_document:
    toc: yes
  html_notebook:
    theme: lumen
    toc: yes
---

# Load Libraries
```{r load library, include=FALSE}

# Installation of some useful packages
suppressPackageStartupMessages({c(
  if(!isTRUE(require("prettydoc"))){install.packages("prettydoc")},
  if(!isTRUE(require("readxl"))){install.packages("readxl")},
  library(prettydoc), # htm template
  library(readxl), # reads excel files
  library(reshape2), # for acast() - transform from long to short format data
  library(tidyverse), library(RNAseqQC),
  library(RUVSeq), library(edgeR), library(DESeq2),
  library(dplyr), library(tidyr), library(ComplexHeatmap), 
  library(RColorBrewer), library(circlize),
  library(ggplot2), library(ggrepel), library(PCAtools),
  library(EnhancedVolcano), library(cowplot),
  library(kableExtra), 
  library(knitr),
  library(biomaRt)
  )})

# Load custom functions 
source("functions.R")

set.seed(42)
```

# Load and transform data
Load counts and biological data from the "data" folder
```{r}
# load counts data
# ctsToDf is a custom function used to transform longform cts table to short form
cts_raw <- ctsToDf("data/3col.tsv") 
colnames(cts_raw) <- gsub("_.*", "", colnames(cts_raw)) 
y <- length(rownames(cts_raw))

head(cts_raw)
```

## Create transcript to gene table via GTF file
Create folder named "rds" first
```{r}
gtf2transcript2gene <- function(url, destfile_gtf, destfile_rds){
  library(rtracklayer)
  if(!file.exists(destfile_gtf))(download.file(url, destfile_gtf))
  gtf <- import(destfile_gtf)  
  colnames(mcols(gtf))
  
  transcript2gene <- gtf %>%
    mcols() %>%
    as.data.frame() %>%
    dplyr::select(transcript_id, gene_id, gene_name, gene_biotype) %>%
    distinct()

  saveRDS(transcript2gene, destfile_rds)
  
  return(transcript2gene)
}

url <- "https://ftp.ensembl.org/pub/release-114/gtf/homo_sapiens/Homo_sapiens.GRCh38.114.gtf.gz"
destfile_gtf <- "data/Homo_sapiens.GRCh38.114.gtf.gz"
destfile_rds <- "rds/transcript2gene_biotype.rds"
gtf2transcript2gene(url, destfile_gtf, destfile_rds)

transcript2gene_biotype <- readRDS("rds/transcript2gene_biotype.rds")

```

## Map transcript ID to gene ID
```{r eval=FALSE, include=FALSE}
cts_trns2gene <- cts_raw %>%
  tibble::rownames_to_column(., "ensembl_transcript_id_version") %>%
  mutate(ensembl_transcript_id_version = gsub("\\..*","", .$ensembl_transcript_id_version)) %>%
  left_join(., transcript2gene_biotype, by = c("ensembl_transcript_id_version"="transcript_id"))%>% 
  filter(gene_biotype == "protein_coding") %>% 
  as.data.frame()

# check that all transcript IDs have gene IDs
cts_trns2gene %>%
    filter(str_detect(ensembl_transcript_id_version, "_")) %>%
    nrow() # this should be zero

# group by geneID, aggregate counts, and attach gene name
cts <- cts_trns2gene %>% 
  mutate(gene_transcript = ifelse(is.na(gene_id) | 
                                    gene_id == "" | 
                                    gene_id == "_", 
                                  paste0("--", ensembl_transcript_id_version), 
                                  paste0(gene_id))) %>% 
  mutate(gene_transcript = gsub("\\..*","", .$gene_transcript)) %>%
  dplyr::select(-c(ensembl_transcript_id_version, 
                   gene_id, gene_name, gene_biotype)) %>% 
  group_by(gene_transcript) %>% 
  summarise_all(., sum) %>%
  tibble::column_to_rownames(., var = "gene_transcript") %>%
  as.data.frame()

saveRDS(cts, "rds/cts_proteinCodingOnly.rds")
```

## Load phenotype data
```{r}
coldata <- read.csv("data/silver_hskm condition_by group.csv", sep = ",", header = TRUE) %>% 
  arrange(sampleid) %>% 
  mutate(Participant_ID = gsub("-[0-9]+$", "", .$sampleid))

rownames(coldata) <- coldata$sampleid
coldata$condition <- factor(coldata$condition,
                            levels = c("PRE","POST", "3HPOST"))
coldata$sampleid <- factor(coldata$sampleid)
coldata$Participant_ID <- factor(coldata$Participant_ID)

saveRDS(coldata, "rds/bulkRNA_muscle_coldata.rds")

all(rownames(coldata) %in% colnames(cts)) 
all(rownames(coldata) == colnames(cts)) 

cat(" Number of transcript IDs: ", y,"\n",
    "Number of protein coding gene IDs: ", length(rownames(cts)))

```

# Detect outlier samples via unsupervised clustering
```{r fig.width=12}
# detect outlier samples - hierarchical clustering
htree <- hclust(dist(t(cts)), method = "average")
groups <- cutree(htree, k=3) # cut tree into clusters

plot(htree, labels(groups))
# draw dendogram with red borders around the clusters
rect.hclust(htree, k=3, border="red")
```


# Comparison of normalisation methods 
The comparison is made using Relative log expression (RLE) plots (Gandolfo and Speed, 2018).
The normalisation methods compared are: Upper quantile normalisation, Trimmed mean of m-values (TMM - edgeR), and Variance Stabilising Transformation (VST - DESeq2).

## Normalisation for the entire data
This comparison is used with the dataset **including** all the samples.
```{r fig.width=11, fig.height=8}
cts.comp.norm <- cts
p_dat.comp.norm <- coldata %>% 
  mutate(RowName = rownames(.))

group <- p_dat.comp.norm$Type
# normalise via TMM (edgeR)
y <- DGEList(counts = cts.comp.norm, group=group)
keep.exprs <- filterByExpr(y, group=group)
d <- y[keep.exprs,, keep.lib.sizes=FALSE]

d <- calcNormFactors(d, method="TMM") # get tmm from edgeR
d = estimateCommonDisp(d, verbose=TRUE)
tmm_human <- d$pseudo.counts

# VSD from DESEq
dds <- DESeqDataSetFromMatrix(countData = cts.comp.norm,
                              colData = p_dat.comp.norm,
                              design = ~ condition)

keep <- rowMeans(counts(dds)) >= 10 
dds <- dds[keep,]
human.vsd <- vst(dds, blind=T) 

dds <- estimateSizeFactors(dds)
human.deseq <- counts(dds, normalized=TRUE)


set1 <- newSeqExpressionSet(as.matrix(cts.comp.norm),
                            phenoData = AnnotatedDataFrame(data=p_dat.comp.norm, 
                            varMetadata = data.frame(labelDescription=c(names(p_dat.comp.norm)))))
# upper quantile normalisation
set_upper <- betweenLaneNormalization(set1, which="upper")

set_tmm <- newSeqExpressionSet(as.matrix(tmm_human),
                            phenoData = AnnotatedDataFrame(data=p_dat.comp.norm, 
                            varMetadata = data.frame(labelDescription=c(names(p_dat.comp.norm)))))

set_vsd <- newSeqExpressionSet(as.matrix(round(assay(human.vsd))),
                            phenoData = AnnotatedDataFrame(data=p_dat.comp.norm, 
                            varMetadata = data.frame(labelDescription=c(names(p_dat.comp.norm)))))

set_mmDeseq <- newSeqExpressionSet(as.matrix(round(human.deseq)),
                            phenoData = AnnotatedDataFrame(data=p_dat.comp.norm, 
                            varMetadata = data.frame(labelDescription=c(names(p_dat.comp.norm)))))

x <- c(rep(1:30, each=2))
colors <- brewer.pal(9, "Set1")
colors <- colorRampPalette(colors)(60)
plotRLE(set1, outline=FALSE, ylim=c(-1.2, 1.9), col=colors[x], xlab="No Normalisation", xaxt = "n")

plotRLE(set_upper, outline=FALSE, ylim=c(-1, 1), col=colors[x], xlab="Upper qunatile", xaxt = "n") 

plotRLE(set_tmm, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="TMM - EdgeR", xaxt = "n")

plotRLE(set_mmDeseq, outline=FALSE, ylim=c(-3, 3), col=colors[x], xlab="Median of Ratios - DESEq", xaxt = "n")

pcas_names <- c(names(p_dat.comp.norm))

pcaplot_list <- list()

for (i in pcas_names) {
  #nudge <- position_nudge(y = 1)
  p <- plotPCA(human.vsd, intgroup=i) +
    #geom_text(aes(label = name), position = nudge) +
    theme_classic()
  pcaplot_list[[i]] <- p
}

pcaplot_list[1:2]
```


# Session information
```{r}
sessionInfo() 
```

