---
title: "Comparison of normalisation methods"
author: "Megan Soria"
date: 2025-07-21
output:
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

This notebook aims to compare normalisation methods for micro-RNAs (miRNAs) extracted from whole skeletal muscle and mitrochodrial fractions from participants of an excercise experiment. Each sample was taken pre, post, and 3hours post exercise. 

```{r include=FALSE}
# load libraries
suppressPackageStartupMessages({c(
library(RUVSeq),
library(edgeR),
library(DESeq2),
library(tidyverse),
library(readxl),
library(RColorBrewer),
library(RNAseqQC),
library(ggplot2),
library(VennDiagram)
)})

set.seed(42)
```

# Load counts data and filter out zero counts
## muscle miRNA
```{r}
muscle_miRNA_cts <- read.csv("data/MUSC_miRNA_RAW_counts.csv", sep=",", row.names = 1) 
names(muscle_miRNA_cts) <- gsub("_.*", "", names(muscle_miRNA_cts))

muscle_miRNA_coldata <- data.frame("sample_ID" = colnames(muscle_miRNA_cts)) %>% 
  mutate(cond = gsub(".*\\.", "", .$sample_ID)) %>% 
  mutate(condition = if_else(cond == "1", "PRE",
                    if_else(cond == "2", "POST", "3HPOST"))) %>% 
  mutate(Participant_ID = gsub("\\..*", "", .$sample_ID))

muscle_miRNA_coldata$condition <- factor(muscle_miRNA_coldata$condition,
                                         levels = c("PRE", "POST", "3HPOST"))
muscle_miRNA_coldata$sample_ID <- factor(muscle_miRNA_coldata$sample_ID)
muscle_miRNA_coldata$Participant_ID <- factor(muscle_miRNA_coldata$Participant_ID)
rownames(muscle_miRNA_coldata) <- muscle_miRNA_coldata$sample_ID

all(rownames(muscle_miRNA_coldata) == colnames(muscle_miRNA_cts)) #check these both come back as TRUE

saveRDS(muscle_miRNA_coldata, "rds/muscle_miRNA_coldata.rds")
saveRDS(muscle_miRNA_cts, "rds/muscle_miRNA_cts.rds")

```

## mito miRNA
```{r}
mito_cts <- read.csv("data/MITO_miRNA_RAW_counts.csv", sep=",", row.names = 1) 
names(mito_cts) <- gsub("_.*", "", names(mito_cts)) 

# rename hmta.12.2.2 to hmta.12.2
names(mito_cts) <- gsub("^((?:[^.]*\\.){2}[^.]*)\\..*$", "\\1", names(mito_cts))

mito_cts <- mito_cts[, order(names(mito_cts))]

mito_coldata <- data.frame("sample_ID" = colnames(mito_cts)) %>% 
  mutate(cond = gsub(".*\\.", "", .$sample_ID)) %>% 
  mutate(condition = if_else(cond == "1", "PRE",
                    if_else(cond == "2", "POST", "3HPOST"))) %>% 
  mutate(Participant_ID = gsub("(\\w+)\\.(\\d+)\\.\\d+", "\\1\\2", .$sample_ID)) 
  

mito_coldata$condition <- factor(mito_coldata$condition,
                                 levels = c("PRE", "POST", "3HPOST"))
mito_coldata$sample_ID <- factor(mito_coldata$sample_ID)
mito_coldata$Participant_ID <- factor(mito_coldata$Participant_ID)
rownames(mito_coldata) <- mito_coldata$sample_ID

all(rownames(mito_coldata) == colnames(mito_cts)) #check these both come back as TRUE
```

## Remove hmtb09.1 and hmtb09.3 samples (duplicates of the hmta09)
```{r}
mito_coldata <- mito_coldata %>% 
  filter(!Participant_ID == "hmtb09")

mito_cts <- mito_cts %>% 
  select(mito_coldata$sample_ID)

all(rownames(mito_coldata) == colnames(mito_cts))
```


```{r}
datalist = list(mito_miRNA_coldata = mito_coldata, 
                mito_miRNA_cts = mito_cts, 
                muscle_miRNA_coldata = muscle_miRNA_coldata, 
                muscle_miRNA_cts = muscle_miRNA_cts)
saveRDS(datalist, "rds/miRNAdatasets_filtered_hmtb09.rds")
```

# PCA Plot and outlier samples for muscle miRNA data
```{r fig.width=9, fig.height=5}
# detect outlier samples - hierarchical clustering
htree <- hclust(dist(t(muscle_miRNA_cts)), method = "average")
groups <- cutree(htree, k=3) # cut tree into clusters

plot(htree, labels(groups))
# draw dendogram with red borders around the clusters
rect.hclust(htree, k=3, border="red")
```

## Uncorrected PCA
```{r fig.width=9, fig.height=5}
# PCA
# use mito.coldata.nf and mito.cts for unfiltered data (with all samples)
dds_pca <- DESeqDataSetFromMatrix(countData = muscle_miRNA_cts,
                              colData = muscle_miRNA_coldata,
                              design = ~ Participant_ID + condition)
keep <- rowSums(counts(dds_pca)) >= 10
dds_pca <- dds_pca[keep,]

dds_pca.vsd <- varianceStabilizingTransformation(dds_pca)  

plot_pca(dds_pca.vsd, PC_x = 1, PC_y = 2, color_by = "condition", 
         point_rel_size = 5)
```

## Corrected PCA (batch = Participant_ID)
```{r fig.width=9, fig.height=5}
library(limma)
mat_corrected <- removeBatchEffect(assay(dds_pca.vsd), batch = muscle_miRNA_coldata$Participant_ID)
# saveRDS(mat_corrected, "rds/mat_corrected_combinedALL.rds")

assay(dds_pca.vsd) <- mat_corrected

p <- plot_pca(dds_pca.vsd, PC_x = 1, PC_y = 2, color_by = "condition",
         point_rel_size = 5, show_plot = F) 

cbPalette <- c("PRE" = "#00048b", "POST" = "#2463ff", "3HPOST" = "#088F8F")

p_plot <- p$plot + 
  scale_color_manual("Condition", values = cbPalette) +
  geom_hline(aes(yintercept = 0), linetype = 2, color = "black") + 
  geom_vline(aes(xintercept = 0), linetype = 2, color = "black") +
  labs(title = "Corrected Muscle miRNA PCA Plot") +
  theme_minimal(base_size = 15) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 15))

p_plot
```

# PCA Plot and outlier samples for mitochondrial miRNA data
```{r fig.width=9, fig.height=5}
# detect outlier samples - hierarchical clustering
htree <- hclust(dist(t(mito_cts)), method = "average")
groups <- cutree(htree, k=3) # cut tree into clusters

plot(htree, labels(groups))
# draw dendogram with red borders around the clusters
rect.hclust(htree, k=3, border="red")
```

## Uncorrected PCA
```{r fig.width=9, fig.height=9}
# PCA
# use mito.coldata.nf and mito.cts for unfiltered data (with all samples)
dds_pca <- DESeqDataSetFromMatrix(countData = mito_cts,
                              colData = mito_coldata,
                              design = ~ Participant_ID + condition)
keep <- rowSums(counts(dds_pca)) >= 10
dds_pca <- dds_pca[keep,]

dds_pca.vsd <- varianceStabilizingTransformation(dds_pca, blind=T)  

p <- plot_pca(dds_pca.vsd, PC_x = 1, PC_y = 2, color_by = "condition",
         point_rel_size = 5, show_plot = F) 

cbPalette <- c("PRE" = "#00048b", "POST" = "#2463ff", "3HPOST" = "#088F8F")

p_plot <- p$plot + 
  scale_color_manual("Condition", values = cbPalette) +
  geom_hline(aes(yintercept = 0), linetype = 2, color = "black") + 
  geom_vline(aes(xintercept = 0), linetype = 2, color = "black") +
  #geom_text(aes(label = sample_ID), vjust = -0.5) +
  labs(title = "Uncorrected Mitochondrial miRNA PCA Plot") +
  theme_minimal(base_size = 15) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 15))

p_plot

```
## Corrected PCA (batch = Participant_ID)
```{r fig.width=9, fig.height=5}
library(limma)
mat_corrected <- removeBatchEffect(assay(dds_pca.vsd), batch = mito_coldata$Participant_ID)

assay(dds_pca.vsd) <- mat_corrected

p <- plot_pca(dds_pca.vsd, PC_x = 1, PC_y = 2, color_by = "condition",
         point_rel_size = 5, show_plot = F) 

cbPalette <- c("PRE" = "#00048b", "POST" = "#2463ff", "3HPOST" = "#088F8F")

p_plot <- p$plot + 
  scale_color_manual("Condition", values = cbPalette) +
  geom_hline(aes(yintercept = 0), linetype = 2, color = "black") + 
  geom_vline(aes(xintercept = 0), linetype = 2, color = "black") +
  #geom_text(aes(label = sample_ID), vjust = -0.5) +
  labs(title = "Corrected Mitochondrial miRNA PCA Plot") +
  theme_minimal(base_size = 15) +
  theme(panel.grid = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5, size = 15))

p_plot
```

# Session Info
```{r session info}
sessionInfo()
```

