---
title: 'Skeletal Muscle DE Analysis'
author: "MSoria"
date: 2023-06-16
output:
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

## Summary and background information
The data used in this analysis is from an exercise intervention study in human participants. Muscle biopsies were taken at 3 time points for each participant: pre-exercise, post-exercise, and 3 hours post-exercise <br> <br>
A whole muscle and a mitochondrial fraction is extracted from each biopsy. This analysis focuses on the differential expression of the whole muscle transcriptome.  

## Load and install packages

```{r include=FALSE}
suppressPackageStartupMessages({c(
  # packages from CRAN
  if(!require("prettydoc")) install.packages("prettydoc"), library(prettydoc),
  if(!require("readxl")) install.packages("readxl"), library(readxl),
  if(!require("dplyr")) install.packages("dplyr"), library(dplyr),
  if(!require("RColorBrewer")) install.packages("RColorBrewer"), library(RColorBrewer),
  if(!require("circlize")) install.packages("circlize"), library(circlize),
  library(ggplot2),
  library(RNAseqQC),
  # packages from Bioconductor
  # run the installation code when needed
  # if (!require("BiocManager", quietly = TRUE)) install.packages("BiocManager")
  #   BiocManager::install(c("DESeq2", "ComplexHeatmap", "EnhancedVolcano", "org.Hs.eg.db")),
  library(DESeq2),
  library(ComplexHeatmap),
  library(EnhancedVolcano),
  library(cowplot),
  library(PCAtools),
  library(reshape2),
  library(tidyverse)
  )})

# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("DESeq2")
  
# load self-authored functions 
source("functions.R")

set.seed(42)
```

***

## Load cts and col data
```{r}
cts <- readRDS("rds/cts_proteinCodingOnly.rds")
coldata <- readRDS("rds/bulkRNA_muscle_coldata.rds")

all(rownames(coldata) %in% colnames(cts)) 
all(rownames(coldata) == colnames(cts)) 
```

## Differential expression analysis
The following differential expression analysis is conducted using the DESeq2 package
```{r DEseq2,echo=T, results='hide', message=FALSE}
#set up deseq2 object
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ Participant_ID + condition)

dds$condition <- relevel(dds$condition, ref = "PRE") 

keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]

dds <- DESeq(dds) #run deseq2

# Jess's orig matrix has 15,198 rows
cat("Before filtering - total gene IDs: ", length(rownames(cts)), "\n",
    "After filtering - total gene IDs: ", dim(dds)[1])
```

#### Pull out DEseq2 normalised read counts

```{r deseq2_norm, message=FALSE, warning=FALSE, results='hide'}
dds <- estimateSizeFactors(dds)
normcounts <- counts(dds, normalized=TRUE)
```

### Extract all possible contrasts 

```{r}
# extract contrasts
hr3POST_vs_PRE <- results(dds, contrast=c("condition","3HPOST","PRE")) %>%
  .[order(.$log2FoldChange, decreasing = TRUE),] %>%
      as.data.frame(.)

POST_vs_PRE <- results(dds, contrast=c("condition","POST","PRE")) %>%
  .[order(.$log2FoldChange, decreasing = TRUE),] %>%
      as.data.frame(.)

hr3POST_vs_POST <- results(dds, contrast=c("condition","3HPOST","POST")) %>%
  .[order(.$log2FoldChange, decreasing = TRUE),] %>%
      as.data.frame(.)

# store all contrasts in a list 
ddsResults_list <- list(
  "condition_3HPOST_vs_PRE" = hr3POST_vs_PRE,
  "condition_POST_vs_PRE" = POST_vs_PRE,
  "condition_3HPOST_vs_POST" = hr3POST_vs_POST
)

ddsResults_list <- diff_expressed(ddsResults_list, padjust = 0.05, L2FC = 1)
names(ddsResults_list)

saveRDS(ddsResults_list, "rds/ddsResults_wholeMusc.rds")

```

## Visualisations

Apply variance stabilising transformation to the DESeq data
```{r fig.width=10, fig.height=6}
# apply variance stabilising transformation to the DESeq data
# this normalisation is required for calculations involving distances (heatmaps, clustering, etc.)
vsd <- vst(dds, blind=T)

#plotPCA(vsd, intgroup=c("condition"))
pca <- plot_pca(vsd, PC_x = 1, PC_y = 2, color_by = "condition", point_rel_size = 4) 
p <- pca(assay(vsd))

s <- screeplot(p, axisLabSize = 18, titleLabSize = 22) 
s

# pdf(file = "figures/wholeMusc_mRNA_pca.pdf",
#     width = 10, # The width of the plot in inches
#     height = 6) # The height of the plot in inches
# pca
# s
# dev.off()

```

```{r }
# add gene names from ENSEMBL Ids
cts_trns2gene <- readRDS("rds/transcript2gene_biotype.rds")
data_genename <- cts_trns2gene[2:3] %>% 
  distinct()
names(data_genename) <- c("ENSEMBL", "SYMBOL")

de_genes <- ddsResults_list$condition_3HPOST_vs_PRE %>%
                            rownames_to_column(var = "ENSEMBL") %>% 
  filter(padj < 0.05) %>% 
  filter(abs(log2FoldChange) >= 1)

signif_genes <- de_genes %>% 
  left_join(data_genename, by = 'ENSEMBL') %>%
  filter(!is.na(padj)) %>% 
#  filter(!SYMBOL == "") %>% 
   mutate(SYMBOL = ifelse(is.na(SYMBOL) |
                                    SYMBOL == "",
                                  paste0(ENSEMBL),
                                  paste0(SYMBOL))) %>%
  # filter most significant ensembl gene / isoform mapping to the same gene symbol
  group_by(SYMBOL) %>%
  slice_min(order_by = padj, n = 1, with_ties = FALSE) %>%  
  ungroup() %>% 
  column_to_rownames(var = "ENSEMBL") %>% 
  as.data.frame() 
signif_genes <- signif_genes[order(signif_genes$padj, decreasing = F),]
                                  
```

### Heatmap
This heatmap shows the changes in gene expression across all time points using condition = 3HRPOST vs PRE. The top 30 up-regulated and down-regulated genes are filtered according to pvalue < 0.05 and |log2 fold change| > 1
```{r}
mat <- assay(vsd)[rownames(signif_genes),]
mat_z <- t(apply(mat, 1, scale))
colnames(mat_z) <- colnames(mat)

# select the top and bottom genes and store log2foldchange and baseMean values
numkeep <- 70
rowskeep <- c(seq(1:numkeep))

# set a pre, post, and 3hour post groupings
sampling_group <- c(rep(c("Pre", "Post", "3hrPost"), times = 3),
                    c("Post", "3hrPost"), 
                    rep(c("Pre", "Post", "3hrPost"), times = 4),
                    c("Post", "3hrPost"), 
                    c("Pre", "Post"),
                    rep(c("Pre", "Post", "3hrPost"), times = 2))
sampling_group_col = c("Pre" = 5, "Post" = 4, "3hrPost" = 6)

dend1 = cluster_within_group(mat_z, sampling_group)

```

Create the heatmaps
```{r fig.width=14.22, fig.height=15}

h1 <- Heatmap(mat_z[rowskeep,], 
              column_title_gp = gpar(fontsize = 23),
              cluster_rows = T, # cluster only columns to compare samples
              column_labels = colnames(mat_z), 
              row_labels = signif_genes$SYMBOL[rowskeep], 
              name="Z-score", 
              column_names_rot = 45,
              cluster_columns = dend1, 
              column_split = 3,
              top_annotation = HeatmapAnnotation(
                Condition = sampling_group,  
                col = list(Condition = sampling_group_col)))

draw(h1, padding = unit(c(2, 10, 2, 2), "mm"))

```


```{r eval=FALSE, fig.height=15, fig.width=14.22, include=FALSE}
pdf(file = "figures/wholeMusc_mRNA_hmap.pdf",
    width = 14.22, # The width of the plot in inches
    height = 15) # The height of the plot in inches
draw(h1, padding = unit(c(2, 10, 2, 2), "mm"))
dev.off()
```

### Volcano plots
```{r fig.width=14.22, fig.height=15}
plots <- list()
for (i in 1:length(ddsResults_list)) {
   
  upGenes <- table(ddsResults_list[[i]]$diffexpressed)["UP"]
  downGenes <- table(ddsResults_list[[i]]$diffexpressed)["DOWN"]
  name1 <- gsub("_", " ", names(ddsResults_list[i]))
  name <- gsub("condition", "", name1)
     
    p <- EnhancedVolcano(ddsResults_list[[i]], x="log2FoldChange", y="padj", lab = NA, 
                pCutoff = 0.05, FCcutoff = 1, title = name, pointSize = 4,
                subtitle = paste0("Up-regulated = ", upGenes, " and Down-regulated = ", downGenes), 
                gridlines.major = FALSE, gridlines.minor = FALSE, axisLabSize = 20, titleLabSize = 20, subtitleLabSize = 18,
                captionLabSize = 0, legendLabSize = 15, legendIconSize = 8, col = c("grey30", "royalblue", "forestgreen", "red2"))
  
  plots[[i]] <- p
}
plot_grid(plotlist = plots, ncol = 2, labels = LETTERS[1:length(plots)], label_size = 18)
```


```{r eval=FALSE, fig.height=15, fig.width=14.22, include=FALSE}
pdf(file = "figures/wholeMusc_mRNA_volcanoPlots.pdf",
    width = 10, # The width of the plot in inches
    height = 10) # The height of the plot in inches
plots[1:3]
dev.off()
```

### Summary of up-regulated and down-regulated genes
```{r }
diffexpr_summary <- list()

for (i in 1:length(ddsResults_list)) {
  
  upGenes <- table(ddsResults_list[[i]]$diffexpressed)["UP"]
  downGenes <- table(ddsResults_list[[i]]$diffexpressed)["DOWN"]
  name <- gsub("_", " ", names(ddsResults_list[i]))
  name2 <- gsub("condition", "", name)
  
  diffexpr_summary[[paste0(i)]]$Stage <- name2
  diffexpr_summary[[paste0(i)]]$Up <- upGenes
  diffexpr_summary[[paste0(i)]]$Down <- downGenes
}

diffexpr_summaryDF <- do.call(rbind.data.frame, diffexpr_summary) %>%
  as.data.frame(.)
diffexpr_summaryDF
```


```{r eval=FALSE, include=FALSE}
write.csv(diffexpr_summaryDF, "outData/wholeMusc_mRNA_DEsummary.csv")
write.csv(ddsResults_list$condition_3HPOST_vs_PRE, "outData/wholeMusc_mRNA_DE_3HPOSTvPRE.csv")
write.csv(ddsResults_list$condition_POST_vs_PRE, "outData/wholeMusc_mRNA_DE_POSTvPRE.csv")
write.csv(ddsResults_list$condition_3HPOST_vs_POST, "outData/wholeMusc_mRNA_DE_3HPOSTvPOST.csv")
```


Session Info
```{r session info} 
sessionInfo() 
```

