---
title: "Pathway Analysis for Whole Skeletal Muscle"
author: "MSoria"
date: '2025-07-21'
output:
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

# Purpose

This markdown file aims to perform a pathway analysis on the differentially expressed genes of an exercise intervention study.

### Install and load prerequisite R-packages

```{r load library, message=FALSE, warning=FALSE, include=FALSE, paged.print=TRUE}
# Installation of some useful packages
suppressPackageStartupMessages({c(
  if(!isTRUE(require("prettydoc"))){install.packages("prettydoc")},
  if(!isTRUE(require("readxl"))){install.packages("readxl")},
  library(prettydoc),
  library(readxl), # reads excel files
  library(here), # finds files based on current working directory
  library(DESeq2),
  library(biomaRt),
  library(tidyverse),
  library(dplyr),
  library(msigdbr),
  library(clusterProfiler),
  library(enrichplot),
  library(pathview)
  )})

# load self-authored functions 
source("functions.R")

set.seed(42)

```

# Load Differentially Expressed Genes
```{r}
ddsResults_list <- readRDS("rds/ddsResults_wholeMusc.rds")

# create clusterProfiler inputs for ORA and GSEA
pre_3h_input <- cProfilr_input(ddsResults_list[[1]])
pre_post_input <- cProfilr_input(ddsResults_list[[2]])
post_3h_input <- cProfilr_input(ddsResults_list[[3]])

```


# Get gene set data 
Using the **msigdbr library** from the [Broad Institute's Molecular Signatures Database](https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp).

```{r echo=TRUE}
# GO-Biological pathways
GO_BP <- msigdbr(species = "Homo sapiens", 
                 collection = "C5", subcollection = "BP") %>% 
  dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame() %>% 
  mutate(gs_name = gsub("GOBP_", "", .$gs_name)) %>% 
  mutate(gs_name = gsub("_", " ", .$gs_name)) %>%  
  as.data.frame()

# Reactome pathways
reactome <- msigdbr(species = "Homo sapiens", 
                 collection = "C2", subcollection = "CP:REACTOME") %>% 
  dplyr::distinct(gs_name, ensembl_gene) %>% 
  mutate(gs_name = gsub("REACTOME_", "", gs_name)) %>% 
  mutate(gs_name = gsub("_", " ", gs_name)) 
```

# Gene Set Enrichment analysis 
## Pre vs 3hr post 
```{r fig.height=7, fig.width=11, warning=FALSE}
options(enrichment_force_universe = TRUE)

preV3h_gseGO <- clusterProfiler::GSEA(
  geneList = pre_3h_input$all_gene_list, 
                     TERM2GENE = GO_BP,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d1_GO <- dotplot(preV3h_gseGO, 
        showCategory=15, split=".sign", 
        x = "NES",
        label_format = 35,
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "GO BP Pathway Analysis of DE Genes (3hr post vs Pre exercise)") +
        theme_bw(base_size = 18) +
        theme(plot.title=element_text(hjust=0.5, size = 20, face = "bold"),
              plot.subtitle=element_text(vjust=2, hjust=0.5, size = 18,face = "bold"),
              axis.text.y = element_text(size = 11)) 

preV3h_react <- clusterProfiler::GSEA(
  geneList = pre_3h_input$all_gene_list, 
                     TERM2GENE = reactome,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d1_react <- dotplot(preV3h_react, showCategory=15, split=".sign", 
        label_format = 35,
        x = "NES",
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "Reactome Pathway Analysis of DE Genes (3hr post vs Pre exercise)") +
        theme_bw(base_size = 18) +
        theme(plot.title=element_text(hjust=0.5, size = 20, face = "bold"),
              plot.subtitle=element_text(vjust=2, hjust=0.5, size = 18,face = "bold"),
              axis.text.y = element_text(size = 11)) 

d1_GO
d1_react
```


```{r eval=FALSE, fig.height=7, fig.width=11, warning=FALSE, include=FALSE}
pdf(file = "figures/wholeMusc_mRNA_3hPostVpre_pathway.pdf",
    width = 11, # The width of the plot in inches
    height = 7) # The height of the plot in inches
d1_GO
d1_react
dev.off()
```

## Pre vs Post 
```{r fig.height=7, fig.width=8, warning=FALSE}
options(enrichment_force_universe = TRUE)

preVpost_gseGO <- clusterProfiler::GSEA(geneList = pre_post_input$all_gene_list, 
                     TERM2GENE = GO_BP,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d2_GO <- dotplot(preVpost_gseGO, showCategory=15, split=".sign", 
        label_format = 30,
        x = "NES",
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "GO BP GSEA of DE Genes (Post vs Pre exercise)") +
        theme_bw(base_size = 15) +
        theme(plot.title=element_text(hjust=0.5, size = 15, face = "bold"),
              axis.text.y = element_text(size = 12)) 

preVpost_react <- clusterProfiler::GSEA(geneList = pre_post_input$all_gene_list, 
                     TERM2GENE = reactome,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d2_react <- dotplot(preVpost_react, showCategory=15, split=".sign", 
        label_format = 30,
        x = "NES",
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "KEGG GSEA of DE Genes (Post vs Pre exercise)") +
        theme_bw(base_size = 15) +
        theme(plot.title=element_text(hjust=0.5, size = 15, face = "bold"),
              axis.text.y = element_text(size = 12)) 

d2_GO
d2_react
```

```{r eval=FALSE, fig.height=10, fig.width=12, warning=FALSE, include=FALSE}
pdf(file = "figures/wholeMusc_mRNA_preVpost_pathway.pdf",
    width = 8, # The width of the plot in inches
    height = 7) # The height of the plot in inches
d2_GO
d2_react
dev.off()

```


## 3h Post vs Pre
```{r fig.height=7, fig.width=11, warning=FALSE}
options(enrichment_force_universe = TRUE)

postV3hpost_gseGO <- clusterProfiler::GSEA(geneList = post_3h_input$all_gene_list, 
                     TERM2GENE = GO_BP,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d3_GO <- dotplot(preVpost_gseGO, showCategory=10, split=".sign", 
        label_format = 30,
        x = "NES",
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "Gene Set Enrichment Analysis of DE Genes (Post vs Pre exercise)")+
        theme_bw(base_size = 18) +
        theme(plot.title=element_text(hjust=0.5, size = 20, face = "bold"),
              plot.subtitle=element_text(vjust=2, hjust=0.5, size = 18,face = "bold"),
              axis.text.y = element_text(size = 12)) 

postV3hpost_react <- clusterProfiler::GSEA(geneList = post_3h_input$all_gene_list, 
                     TERM2GENE = reactome,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)

# plot results
d3_react <- dotplot(postV3hpost_react, showCategory=10, split=".sign", 
        label_format = 30,
        x = "NES",
        font.size = 5,) +
        facet_grid(.~.sign) + 
        labs(title = "Gene Set Enrichment Analysis of DE Genes (Post vs Pre exercise)")+
        theme_bw(base_size = 18) +
        theme(plot.title=element_text(hjust=0.5, size = 20, face = "bold"),
              plot.subtitle=element_text(vjust=2, hjust=0.5, size = 18,face = "bold"),
              axis.text.y = element_text(size = 12)) 
d3_GO
d3_react
```


```{r eval=FALSE, fig.height=10, fig.width=12, warning=FALSE, include=FALSE}
pdf(file = "figures/wholeMusc_mRNA_preVpost_pathway.pdf",
    width = 11, # The width of the plot in inches
    height = 7) # The height of the plot in inches
d3_GO
d3_react
dev.off()
```

# Compare gene lists to each other

```{r fig.width=11, message=FALSE, warning=FALSE}
# put all gene lists for each DE contrast in one list
compare_list <- list(postVsPre = pre_post_input$all_gene_list,
                     post3hrVsPre = pre_3h_input$all_gene_list)

ck_GO <- compareCluster(geneClusters=compare_list, fun = "GSEA", 
                     TERM2GENE = GO_BP,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)

ck_react <- compareCluster(geneClusters=compare_list, fun = "GSEA", 
                     TERM2GENE = reactome,
                     pvalueCutoff = 0.05,
                     minGSSize = 10, 
                     maxGSSize = 800,
                     eps = 0,
                     by = "fgsea",
                     verbose = FALSE)


compare_GO <- dotplot(ck_GO, showCategory=10, split=".sign",
        label_format = 40,
        font.size = 10,) +
        facet_grid(.~.sign) + 
        labs(title = "GO BP pathway comparison 
             for PreVsPost and PreVs3hPost DE Genes") +
        theme(plot.title=element_text(hjust=0.5, size = 18, face = "bold"),
              plot.subtitle=element_text(hjust=0.5, size = 15, face = "bold"))

compare_react <- dotplot(ck_react, showCategory=10, split=".sign",
        label_format = 40,
        font.size = 10,) +
        facet_grid(.~.sign) + 
        labs(title = "Reactome pathway comparison 
             for PreVsPost and PreVs3hPost DE Genes") +
        theme(plot.title=element_text(hjust=0.5, size = 18, face = "bold"),
              plot.subtitle=element_text(hjust=0.5, size = 15, face = "bold"))

compare_GO 
compare_react
```

Session Info
```{r session info} 
sessionInfo() 
```

