---
title: "mRNA and miRNA expression profile analysis"
author: "Megan Soria"
date: 2025-07-21 
output:
  html_document:
    theme: lumen
    toc: yes
    toc_float: true
  html_notebook:
    theme: lumen
    toc: yes
    toc_float: true
  prettydoc::html_pretty:
    theme: hpstr
    highlight: github
---

This notebook aims to consolidate mRNA and miRNA expression profiles form from 12 participants of an exercise experiment. Each sample was taken pre, post, and 3hours post exercise. This analysis mainly uses the `mirTarRnaSeq` package (Movassagh et al., 2022)

```{r include=FALSE}
#load libraries
library(mirTarRnaSeq)
library(tidyverse)
library(dplyr)
library(biomaRt)
library(fgsea)
library(msigdbr)
library(clusterProfiler)
library(cowplot)

source("functions.R")
set.seed(42)
```
 
# Muscle mRNA and miRNA  correlations across 3 time points  
```{r fig.height=11, fig.width=4.5}
# Load and transform mRNA and miRNA DESeq results
mrna_files <- readRDS("rds/ddsResults_wholeMusc.rds")
mrna_files <- list(c3HPOST_vs_PRE = mrna_files$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj)) %>% 
                     rename(pvalue = padj),
                   cPOST_vs_PRE = mrna_files$condition_POST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj),
                   c3HPOST_vs_POST = mrna_files$condition_3HPOST_vs_POST %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj)
                   )
# Combine into one dataset
mrna <- one2OneRnaMiRNA(mrna_files, 
                        gene_colname = "gene",
                        fc_colname ="log2FoldChange", 
                        pthreshold = 0.05)$foldchanges

mirna_files <- readRDS("rds/ddsResults_musc_miR.rds")
mirna_musc <- list(c3HPOST_vs_PRE = mirna_files$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)),
                   cPOST_vs_PRE = mirna_files$condition_POST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)),
                   c3HPOST_vs_POST = mirna_files$condition_3HPOST_vs_POST %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)))
mirna <- one2OneRnaMiRNA(mirna_musc, 
                         gene_colname = "miR_ID",
                         fc_colname ="log2FoldChange")$foldchanges

corr_0_musc <- corMirnaRna(mrna, mirna, method="pearson")

# Make a background distribution correlation
outs_musc <- sampCorRnaMirna(mrna, mirna,method="pearson",
Shrounds = 100, Srounds = 1000)

#Identify significant correlation
sig_corrs_musc <- threshSig(corr_0_musc, outs_musc, pvalue = 0.05)

#Import concordant miRanda file
# output columns: miRanda score, folding energy, seed match score, and alignment characteristics score
# threshold = 169 (score greater than the 99th percentile of all miRanda interaction scores (miRanda score > 168)) according to paper
miRanda <- getInputSpecies("Human1", threshold = 169)
saveRDS(miRanda, "rds/miRanda_2025.rds")

# miRanda <- readRDS("rds/miRanda_2025.rds")

#Extract your target correlations based on miRanda and correlation threshold. 
newcorr_musc <- corMirnaRnaMiranda(mrna, mirna, -0.7, miRanda)
saveRDS(newcorr_musc, "rds/miMusc_mRNAmusc_corrMiranda.rds")

#Make final results file for significant correlations intersecting with miRanda file
results_musc <- miRandaIntersect(sig_corrs_musc, outs_musc, mrna, mirna, miRanda)
saveRDS(results_musc, "rds/miMusc_mRNAmusc_results.rds")
 
```

## Overrepresentation analysis of significant gene targets with miRanda intersection
```{r fig.width=7, fig.height=9}
# GO-Biological pathways
GO_BP <- msigdbr(species = "Homo sapiens", 
                 collection = "C5", subcollection = "BP") %>% 
  dplyr::distinct(gs_name, ensembl_gene) %>% as.data.frame()
msigdbr_list = split(x = GO_BP$ensembl_gene, f = GO_BP$gs_name)

signif_genes <- unique(results_musc$corrs$V2)
univ <- unique(mrna_files$c3HPOST_vs_PRE$gene)

muscMirna_fora <- fora(msigdbr_list, signif_genes, univ)
saveRDS(muscMirna_fora, "rds/muscMirna_fora.rds")

muscMirna_fora_sig <- muscMirna_fora %>% 
  mutate(Enrich.Score = -log10(pval)) %>% 
  filter(pval <= 0.05) %>% 
  arrange(desc(pval)) %>%
    mutate(pathway = gsub("GOBP_","", pathway),
           pathway = gsub("_"," ", pathway)) %>% 
  top_n(30)
#write.csv(muscMirna_fora_sig[, -6], "outData_round2/muscMirna_fora_sig.csv")

x <- ggplot(muscMirna_fora_sig, aes(x=reorder(pathway, pval),
    y=Enrich.Score, fill=pval)) + geom_col(stat="identity") +
    coord_flip() + scale_fill_gradient(low="darkblue",high="red") +
    labs(x="Pathway", y= paste0("-log10(pval)"),
         title = "Top 30 Enriched GO Biological Pathways",
         subtitle = str_wrap("from significant muscle mRNA and miRNA fold change correlation with miRanda intersection for all time points", width = 70)) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 33)) +
  theme_linedraw(base_size = 10.5) +
    theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 13),
          axis.title.y = element_text(size = 16))

# pdf(file = "figures/miMusc_mRNA_sigCorrMiranda_oraGO.pdf",
#     width = 7, # The width of the plot in inches
#     height = 9) # The height of the plot in inches
# x
# dev.off()

x
```

------

# Muscle mRNA and mitochondrial miRNA  correlations across 3 time points
```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, results='hide'}
mrna_files <- readRDS("rds/ddsResults_wholeMusc.rds")

mrna_files <- list(c3HPOST_vs_PRE = mrna_files$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj)) %>% 
                     rename(pvalue = padj),
                   cPOST_vs_PRE = mrna_files$condition_POST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj),
                   c3HPOST_vs_POST = mrna_files$condition_3HPOST_vs_POST %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj)
                   )

mrna <- one2OneRnaMiRNA(mrna_files, 
                        gene_colname = "gene",
                         fc_colname ="log2FoldChange",
                        pthreshold = 0.05)$foldchanges

# mirna_files <- readRDS("Meg/rds/dddsResults_miRmito.rds")
mirna_files <- readRDS("rds/ddsResults_mito_miR.rds")
mirna_mito <- list(c3HPOST_vs_PRE = mirna_files$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)),
                   cPOST_vs_PRE = mirna_files$condition_POST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)),
                   c3HPOST_vs_POST = mirna_files$condition_3HPOST_vs_POST %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)))
mirna <- one2OneRnaMiRNA(mirna_mito, gene_colname = "miR_ID",
                         fc_colname ="log2FoldChange")$foldchanges

# Make a background distribution correlation
outs_mito <- sampCorRnaMirna(mrna, mirna,method="pearson",
Shrounds = 100)

corr_0_mito <- corMirnaRna(mrna, mirna, method="pearson")

#Identify significant correlation
sig_corrs_mito <- threshSig(corr_0_mito, outs_mito, pvalue = 0.05)

 #Extract your target correlations based on miRanda and correlation threshold.
# output columns: miRanda score, folding energy, seed match score, and alignment characteristics score
#miRanda <- getInputSpecies("Human1", threshold = 168)
#saveRDS(miRanda, "rds/miRanda.rds")

newcorr_mito <- corMirnaRnaMiranda(mrna, mirna, -0.7, miRanda)
saveRDS(newcorr_mito, "rds/miMito_mRNAmusc_corrMiranda.rds")

#Make final results file for significant correlations intersecting with miRanda file
results_mito <- miRandaIntersect(sig_corrs_mito, outs_mito, mrna, mirna, miRanda)
saveRDS(results_mito, "rds/miMito_mRNAmusc_results.rds")

```

## Plots
```{r fig.height=10, fig.width=15}
library(pheatmap)
xx <- mirRnaHeatmap(results_mito$corrs, #upper_bound =-0.99,
              main = str_wrap("Target correlations based on miRanda and correlation threshold (>=0.9) for all time points)", width = 50))

xx2 <- mirRnaHeatmap(results_musc$corrs, #upper_bound =-0.99,
              main = str_wrap("Target correlations based on miRanda and correlation threshold (>=0.9) for all time points)", width = 50))

# save_pheatmap_pdf(xx, filename = "figures/miMito_mRNAmusc_sigCorrMirandaHmap.pdf", width = 35, height = 20)

grid::grid.newpage()
grid::grid.draw(xx$gtable)
```

## Overrepresentation analysis of significant gene targets with miRanda intersection
```{r fig.width=7, fig.height=10}
signif_genes <- unique(results_mito$corrs$V2)
univ <- unique(mrna_files$c3HPOST_vs_PRE$gene)

mitoMirna_fora <- fora(msigdbr_list, signif_genes, univ)

mitoMirna_fora_sig <- mitoMirna_fora %>% 
  mutate(Enrich.Score = -log10(pval)) %>% 
  filter(pval <= 0.05) %>% 
  arrange(desc(pval)) %>%
    mutate(pathway = gsub("GOBP_","", pathway),
           pathway = gsub("_"," ", pathway)) %>% 
  top_n(30)
#write.csv(mitoMirna_fora_sig[, -6], "outData_round2/mitoMirna_fora_sig.csv")

x_0 <- ggplot(mitoMirna_fora_sig, aes(x=reorder(pathway, pval),
    y=Enrich.Score, fill=pval)) + geom_col(stat="identity") +
    coord_flip() + scale_fill_gradient(low="darkblue",high="red") +
    labs(x="Pathway", y= paste0("-log10(pval)"),
         title = "Top 30 Enriched GO Biological Pathways",
         subtitle = str_wrap("from significant muscle mRNA and mitochondrial miRNA fold change correlation with miRanda intersection for all time points", width = 70)) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 35)) +
  theme_linedraw(base_size = 10.3) +
    theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 13),
          axis.title.y = element_text(size = 16))
x_0
# pdf(file = "figures/miMito_mRNA_sigCorrMiranda_oraGO.pdf",
#     width = 7, # The width of the plot in inches
#     height = 9) # The height of the plot in inches
# x_0
# dev.off()
```

-----

# Muscle mRNA and mitochondrial miRNA correlations at 3hr post vs pre exercise
```{r}
mrna_file <- readRDS("rds/ddsResults_wholeMusc.rds")
mrna_file1 <- list(c3HPOST_vs_PRE = mrna_file$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj)
                   )

mrna <- one2OneRnaMiRNA(mrna_file1, 
                        gene_colname = "gene",
                         fc_colname ="log2FoldChange",
                        pthreshold = 0.05)$foldchanges

mirna_file2 <- readRDS("rds/ddsResults_mito_miR.rds")
mirna_file2 <- list(c3HPOST_vs_PRE.musc = mirna_file2$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)))
mirna <- one2OneRnaMiRNA(mirna_file2, 
                         gene_colname = "miR_ID",
                         fc_colname ="log2FoldChange")$foldchanges

# Estimate the miRNA mRNA FC differences for your dataset
inter0 <- twoTimePoint(mrna, mirna)

#Make a background distribution for your miRNA mRNA FC differences
outs <- twoTimePointSamp(mrna, mirna, Shrounds = 100 ) 

#Identify miRNA mRNA relationships bellow a P value threshold, default is 0.05
sig_InterR <- threshSigInter(inter0, outs, pvalue = 0.05)

#Intersect the mirRanda file with your output results
results <- mirandaIntersectInter(sig_InterR, outs, mrna, mirna, miRanda)

#Create a results file for heatmap
final_results_mito <- finInterResult(results)
#Draw plots of miRNA mRNA fold changes for your results file
par(mar=c(4,4,2,1))
drawInterPlots(mrna,mirna,final_results)

CorRes_mito<-results$corrs
```

### Plots
```{r fig.height=10, fig.width=10}

mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
attr <- listAttributes(mart)
genes <-  CorRes_mito$V2
gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","external_gene_name"),
              values = genes, mart= mart) %>% 
  rename("V2" = "ensembl_gene_id")

gene_IDs <- gene_IDs %>%
mutate(external_gene_name = na_if(external_gene_name, "")) %>%
  mutate(external_gene_name = coalesce(external_gene_name, V2))

resTable_mito <- final_results_mito %>% 
  mutate(mRNA = gsub("\\..*","", .$mRNA),
         miRNA = gsub("\\..*","", .$miRNA)) %>% 
  left_join(., CorRes_mito[,c("V2", "V3", "V6")], by = c("mRNA"="V2")) %>% 
  left_join(., gene_IDs, by = c("mRNA"="V2")) %>% 
  distinct() %>% 
  rename(miRandaScore = V3) %>% 
  rename(AlignmentRange = V6)

CorResName_mito <- CorRes_mito %>% 
  left_join(., gene_IDs, by = "V2") %>% 
  dplyr::select(-V2) %>% 
  rename("V2" = "external_gene_name") %>% 
  relocate(V2, .after = V1) %>% 
  distinct()

x_1 <- mirRnaHeatmapDiff(CorResName_mito,
              main = str_wrap("miRanda intersection with mussle mRNA and mitochondrial miRNA correlations at 3hr post vs pre exercise (pval = 0.05)", width =50))

plot_pheatmap(x_1)

# save_pheatmap_pdf(x_1, filename = "figures/3hrVSpre_miMito_mRNAmusc_sigCorrMirandaHmap_pval0.05.pdf", width = 10, height = 10)
```

# ORA for muscle mRNA and mito miRNA
```{r fig.width=7, fig.height=12}
signif_genes <- unique(CorRes_mito$V2)
univ <- unique(mrna_files$c3HPOST_vs_PRE$gene)

mitoMirna_fora_2time <- fora(msigdbr_list, signif_genes, univ)

mitoMirna_fora_2time_sig <- mitoMirna_fora_2time %>% 
  mutate(Enrich.Score = -log10(pval)) %>% 
  filter(pval < 0.05) %>% 
  arrange(desc(pval)) %>%
    mutate(pathway = gsub("GOBP_","", pathway),
           pathway = gsub("_"," ", pathway)) %>% 
  top_n(30)

x_2 <- ggplot(mitoMirna_fora_2time_sig, aes(x=reorder(pathway, pval),
    y=Enrich.Score, fill=pval)) + geom_col(stat="identity") +
    coord_flip() + scale_fill_gradient(low="darkblue",high="red") +
    labs(x="Pathway", y= paste0("-log10(pval)"),
         title = "Top 30 Enriched GO Biological Pathways",
         subtitle = str_wrap("from significant muscle mRNA and mitochondrial miRNA fold change correlation with miRanda intersection (3hr post vs pre-exercise)", width = 70)) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 40)) +
  theme_linedraw(base_size = 10.5) +
    theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 13),
          axis.title.y = element_text(size = 16))

pdf(file = "figures/miMito_mRNA_sigCorrMiranda2time_oraGO_pVAL0.05.pdf",
    width = 7, # The width of the plot in inches
    height = 12) # The height of the plot in inches
x_2
dev.off()

x_2
```

----

# Muscle mRNA and miRNA correlations at 3hr post vs pre exercise
### Calculations (pre-filtered/significant DEgenes only for whole muscle mRNA)
```{r }
mrna_file <- readRDS("rds/ddsResults_wholeMusc.rds")
mrna_file1 <- list(c3HPOST_vs_PRE = mrna_file$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "gene") %>%
                     mutate(gene = gsub("\\..*","", .$gene)) %>%
                     dplyr::select(c(gene, log2FoldChange, padj))%>% 
                     rename(pvalue = padj)
                   )

mrna <- one2OneRnaMiRNA(mrna_file1, 
                        gene_colname = "gene",
                         fc_colname ="log2FoldChange",
                        pthreshold = 0.05)$foldchanges

mirna_file2 <- readRDS("rds/ddsResults_musc_miR.rds")
mirna_file2 <- list(c3HPOST_vs_PRE.musc = mirna_file2$condition_3HPOST_vs_PRE %>%
                     rownames_to_column(., var = "miR_ID") %>%
                     mutate(miR_ID = gsub("\\..*","", .$miR_ID)) %>%
                     dplyr::select(c(miR_ID, log2FoldChange, pvalue)))
mirna <- one2OneRnaMiRNA(mirna_file2, 
                         gene_colname = "miR_ID",
                         fc_colname ="log2FoldChange")$foldchanges


# Estimate the miRNA mRNA FC differences for your dataset
inter0 <- twoTimePoint(mrna, mirna)

#Make a background distribution for your miRNA mRNA FC differences
outs <- twoTimePointSamp(mrna, mirna,Shrounds = 100)

#Identify miRNA mRNA relationships bellow a P value threshold, default is 0.05
sig_InterR <- threshSigInter(inter0, outs, pvalue = 0.05)

#Intersect the mirRanda file with your output results
results <- mirandaIntersectInter(sig_InterR, outs, mrna, mirna, miRanda)

#Create a results file for heatmap
final_results_musc <- finInterResult(results) 
#Draw plots of miRNA mRNA fold changes for your results file
par(mar=c(4,4,2,1))
drawInterPlots(mrna,mirna,final_results)

CorRes_musc<-results$corrs
```

### Plots
```{r fig.height=14, fig.width=12}
mart <- useDataset("hsapiens_gene_ensembl", useMart("ensembl"))
attr <- listAttributes(mart)
genes <-  CorRes_musc$V2
gene_IDs <- getBM(filters= "ensembl_gene_id", attributes= c("ensembl_gene_id","external_gene_name"),
              values = genes, mart= mart) %>% 
  rename("V2" = "ensembl_gene_id")

gene_IDs <- gene_IDs %>%
mutate(external_gene_name = na_if(external_gene_name, "")) %>%
  mutate(external_gene_name = coalesce(external_gene_name, V2))

resTable_musc <- final_results_musc %>% 
  mutate(mRNA = gsub("\\..*","", .$mRNA),
         miRNA = gsub("\\..*","", .$miRNA)) %>% 
  left_join(., CorRes_musc[,c("V2", "V3", "V6")], by = c("mRNA"="V2")) %>% 
  left_join(., gene_IDs, by = c("mRNA"="V2")) %>% 
  distinct() %>% 
  rename(miRandaScore = V3) %>% 
  rename(AlignmentRange = V6)

write.csv(resTable_musc, "outData_round2/resTable_musc_pval0.05.csv")

CorResName_musc <- CorRes_musc %>% 
  left_join(., gene_IDs, by = "V2") %>% 
  dplyr::select(-V2) %>% 
  rename("V2" = "external_gene_name") %>% 
  relocate(V2, .after = V1)

x_3 <- mirRnaHeatmapDiff(CorResName_musc,#upper_bound = 9.9,
              main = str_wrap("miRanda intersection with mussle mRNA and miRNA correlations at 3hr post vs pre exercise (pval = 0.05)", width =50)) 

plot_pheatmap(x_3)

save_pheatmap_pdf(x_3, filename = "figures/3hrVSpre_miMusc_mRNAmusc_sigCorrMirandaHmap_pval0.05.pdf", width = 12, height = 14)

```

# ORA for muscle mRNA and miRNA
```{r fig.width=7, fig.height=9}
signif_genes <- unique(CorRes_musc$V2)
univ <- unique(mrna_files$c3HPOST_vs_PRE$gene)

muscMirna_fora_2time <- fora(msigdbr_list, signif_genes, univ)

muscMirna_fora_2time_sig <- muscMirna_fora_2time %>% 
  mutate(Enrich.Score = -log10(pval)) %>% 
  filter(pval <= 0.05) %>% 
  arrange(desc(pval)) %>%
    mutate(pathway = gsub("GOBP_","", pathway),
           pathway = gsub("_"," ", pathway)) %>% 
  top_n(30)

x_4 <- ggplot(muscMirna_fora_2time_sig, aes(x=reorder(pathway, pval),
    y=Enrich.Score, fill=pval)) + geom_col(stat="identity") +
    coord_flip() + scale_fill_gradient(low="darkblue",high="red") +
    labs(x="Pathway", y= paste0("-log10(pval)"),
         title = "Top 30 Enriched GO Biological Pathways",
         subtitle = str_wrap("from significant muscle mRNA and miRNA fold change correlation with miRanda intersection (3hr post vs pre-exercise)", width = 70)) + 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 33)) +
  theme_linedraw(base_size = 10.5) +
    theme(plot.title=element_text(hjust=0.5),
          plot.subtitle=element_text(hjust=0.5),
          panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 13),
          axis.title.y = element_text(size = 16))

pdf(file = "figures/miMusc_mRNA_sigCorrMiranda2time_oraGO_pVAL0.05.pdf",
    width = 7, # The width of the plot in inches
    height = 9) # The height of the plot in inches
x_4
dev.off()

x_4
```

# ResTable classification
```{r}
path1 <- "outData_round2/resTable_musc_pval0.05.csv"
path2 <- "outData_round2/resTable_mito_pval0.05.csv"

restable <- read.csv(path2) %>% 
  select(-X) %>% 
  group_by(mRNA, miRNA) %>%
  arrange(pvalue, desc(miRandaScore), .by_group = TRUE) %>%
  slice(1) %>%
  ungroup() %>% 
  mutate(Relationship = case_when(
    # miRNA significantly Up, mRNA significantly Down
    FC_miRNA > 0 & FC_mRNA < 0 ~ "Negative Regulation",
    
    # miRNA significantly Down, mRNA significantly Up
    FC_miRNA < 0 & FC_mRNA > 0 ~ "Inverse Regulation",
    
    # Both miRNA and mRNA significantly Up
    FC_miRNA > 0 & FC_mRNA > 0 ~ "Coregulation (increase)",
    
    # Both miRNA and mRNA significantly Down
    FC_miRNA < 0 & FC_mRNA < 0 ~ "Coregulation (decrease)",
    
    # Fallback for any cases not meeting thresholds (if applicable)
    TRUE ~ "Other/NS"
  ) 
  )
write.csv(restable, path2)
```

## Common enriched pathways
```{r fig.height=110, fig.width=12}
common_paths <- inner_join(muscMirna_fora_2time, mitoMirna_fora_2time, by = "pathway") %>% 
  filter(pval.x <= 0.05, pval.y <= 0.05) %>% 
    mutate(pathway = gsub("GOBP_","", pathway),
           pathway = gsub("_"," ", pathway),
           Enrich.Score.x = -log10(pval.x),
           Enrich.Score.y = -log10(pval.y)) %>% 
  relocate(Enrich.Score.x, .after = overlapGenes.x)

write.csv(common_paths, "outData_round2/common_paths.csv")

common_pathsLong <- gather(common_paths, key="measure", value="value", names(common_paths)[-1]) %>% filter(measure == c("Enrich.Score.x", "Enrich.Score.y"))

x_5 <- ggplot(common_paths[, 1:8], aes(x=pathway,
    y=Enrich.Score.x, fill=pval.x)) + geom_col(stat="identity") +
    coord_flip() + 
  scale_fill_gradient(low="darkblue",high="red", name = "Muscle p-value") +
    labs(x="Pathway", y= paste0("-log10(pval)"))+ 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 45)) +
  theme_linedraw(base_size = 16) +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 18),
          axis.title.y = element_text(size = 30),
          legend.position="top",
          legend.key.width = unit(0.6, 'in'),
          legend.title = element_text(size=20))
```


```{r fig.height=17, fig.width=12}
x_6 <- ggplot(common_paths[, c(1, 9:15)], aes(x=pathway,
    y=Enrich.Score.y, fill=pval.y)) + geom_col(stat="identity") +
    coord_flip() + 
  scale_fill_gradient(low="darkblue",high="red", name = "Mito p-value") +
    labs(x="Pathway", y= paste0("-log10(pval)"))+ 
  scale_x_discrete(labels = function(x) str_wrap(x, width = 35)) +
  theme_linedraw(base_size = 16) +
    theme(panel.grid.major = element_blank(), 
          panel.grid.minor = element_blank(),
          axis.title.x = element_text(size = 18),
          axis.title.y = element_blank(),
          axis.text.y = element_blank(),
          legend.position="top",
          legend.key.width = unit(0.6, 'in'),
          legend.title = element_text(size=20))
```


```{r fig.height=125, fig.width=20}
z <- cowplot::plot_grid(x_5, x_6, ncol = 2, labels = c("Common pathways (3hr post vs pre-exercise)", ""), rel_widths = c(1, 0.6))
pdf(file = "figures/miMito_mRNA_commonPaths2time.pdf",
    width = 20, # The width of the plot in inches
    height = 135) # The height of the plot in inches
z
dev.off()
z
```


### References
Movassagh, M. et al. (2022) ‘mirTarRnaSeq: An R/Bioconductor Statistical Package for miRNA-mRNA Target Identification and Interaction Analysis’, BMC Genomics, 23(1), pp. 1–17. Available at: https://doi.org/10.1186/s12864-022-08558-w.

# Session Info
```{r session info}
sessionInfo()
```

